# The Geometry of Vectors

In this chapter you will be introduced to the geometry of vectors and vector operations. It is important to recognize that although the illustration of these ideas is restricted to 2-dimensional space, all of these ideas can be extended to *n*-dimensions. 



```{r create_graph_paper, echo=FALSE}
x = data.frame(x = -15:15) 
y = data.frame(y = -15:15)
fake_graph = tidyr::crossing(x, y)

graph_paper = ggplot(data = fake_graph, aes(x = x, y = y)) +
  geom_point(alpha = 0) +
  theme_light() +
  scale_x_continuous(name ="", breaks = -15:15, labels = NULL, expand = c(0, 0)) +
  scale_y_continuous(name = "", breaks = -15:15, labels = NULL, expand = c(0, 0)) +
  geom_segment(
    x = 0, xend = 0, y = 15, yend = -15, arrow = arrow(length = unit(0.02, "npc"), end = "both", type = "closed")
    ) +
  geom_segment(
    x = -15, xend = 15, y = 0, yend = 0, arrow = arrow(length = unit(0.02, "npc"), end = "both", type = "closed")
    ) +
  theme(
    panel.grid = element_line(color = "#60CDE4"),
    panel.grid.minor = element_blank()
  )
```

<br />


## Vectors

A vector can be represented geometrically by using a coordinate system where each element in the vector corresponds to a distance along one of the reference axes defining the coordinate system. Consider the column vector **a**:

$$
\mathbf{a} = \begin{bmatrix}
2 \\ 3
\end{bmatrix}
$$

This vector has a distance of 2 on the first reference axis (R1) and a distance of 3 on the second reference axis (R2). 

```{r echo=FALSE, fig.width=6, fig.height = 6, out.width='60%', fig.cap="Plot showing vector **a** (in red) in the R1--R2 dimensional space."}
graph_paper +
  geom_segment(x = 0, xend = 2, y = 0, yend = 3, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4) +
  annotate("text", x = 2.25, y = 3.25, label = "(2, 3)", size = 4, hjust = 0, vjust = 0, color = "#FF2D21")
```

Here the reference coordinate system (R1--R2) define a two-dimensional space. Vector **a** had two elements and thus resides in a two-dimensional space. In general, an *n*-dimensional vector resides in an *n*-dimensional space. 

The reference axes always intersect at the origin of the space (i.e., the reference point $(0, 0)$). Here the reference axes are at right angles to one another, although this is not a requirement. When the reference axes are at right angles to each other, the reference system is referred to as an *orthogonal coordinate system*.

<br />


## Vector Properties

Geometrically, a vector is displayed as a line segment with an arrowhead at one end of the segment. The end of the vector with the arrowhead is called the *head* or *terminus*, and the other end of the vector is called the *tail* or *origin*. The head of the vector indicates the direction of the vector. 

All vectors have a direction. They also have a length. These two properties completely define a vector within a given reference system. Note that location of the vector in the reference system is not a property of the vector. For example, the red vector and the blue vector below are identical---they have the same direction and length!

```{r echo=FALSE, fig.width=6, fig.height = 6, out.width='60%', fig.cap="Plot showing two identical vectors (in red and blue) in the R1--R2 dimensional space. They are identical because they have the same length and direction. Location in the reference system is a convenience, not a property of the vector."}
graph_paper +
  # Red vector
  geom_segment(x = 0, xend = 2, y = 0, yend = 3, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  # Blue vector
  geom_segment(x = 5, xend = 7, y = -7, yend = -4, color = "#1A5C71",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4)
```

Location is a convenience in that we can re-locate any vector within the reference system to make it easier to work with. 

<br />


### Vector Length

The length of a vector (also referred to as the vector's *norm*) is the distance from the vector's tip to tail. In an orthogonal reference system, if we locate the tail of the vector at the origin, multiple right triangles with the vector as hypotenuse can be formed with the reference axes. The [Pythagorean Theorem](https://en.wikipedia.org/wiki/Pythagorean_theorem) can then be applied to determine the length of a vector.

```{r fig03, echo=FALSE, fig.width=6, fig.height = 6, out.width='80%', fig.cap="Plot showing vector **a** (in red) in the R1--R2 dimensional space. This vector is the hypotenuse of a right triangle with legs of length 2 and 3."}

triangle_data = data.frame(x = c(0, 2, 2), y = c(0, 0, 3))

graph_paper +
  #geom_segment(x = 0, xend = 2, y = 0, yend = 0, color = "#FF2D21", linetype = "dashed") +
  #geom_segment(x = 0, xend = 0, y = 0, yend = 3, color = "#FF2D21", linetype = "dashed") +
  geom_polygon(data = triangle_data, aes(x = x, y = y), fill = "#FF2D21", alpha = 0.2) +
  geom_segment(x = 0, xend = 2, y = 0, yend = 3, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  ggbrace::geom_brace(xstart = 0, xend = 2, ystart = 0, yend = -0.5, color = "#48556C", rotate = 180) +
  ggbrace::geom_brace(xstart = 2, xend = 2.5, ystart = 0, yend = 3, color = "#48556C", rotate = 90) +
  coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5)) +
  annotate("text", x = 4.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 4.5, label = "R2", size = 4) +
  annotate("text", x = 1, y = -1.25, label = "2", size = 3, vjust = 0, color = "#48556C") +
  annotate("text", x = 2.75, y = 1.5, label = "3", size = 3, hjust = 0, color = "#48556C") +
  annotate("text", x = 0.65, y = 1.85, label = "sqrt(13)", size = 3, color = "#48556C", parse = TRUE, angle = 56.3) 
```

In our example, vector **a** is the hypotenuse of a right triangle with legs of length 2 and 3. We can use the Pythagorean Theorem to find the length of the vector:

$$
\begin{split}
C^2 &= A^2 + B^2 \\[2ex]
&= 2^2 + 3 ^2 \\[2ex]
&= 13 \\[2ex]
C &= \sqrt{13}
\end{split}
$$

Note that the sum of the squared elements is equal to the square of the length of the vector. Then to get the length of **a** (denoted as $\lvert\lvert \mathbf{a} \rvert\rvert$), we take the square root of that sum. For our two-dimensional vector, this is

$$
\lvert\lvert \mathbf{a} \rvert\rvert = \sqrt{a_1^2 + a_2^2}
$$
We can generalize this to finding the length of a vector **a** with *n*-dimensions:

$$
\lvert\lvert \mathbf{a} \rvert\rvert = \sqrt{a_1^2 + a_2^2 + a_3^2 + \ldots + a_n^2}
$$

Since the sum under the square root is equivalent to  computing the inner product of **a** with itself, the length of an *n*-dimensional vector **a**  can be found by computing the square root of the inner product between **a** and itself:

$$
\lvert\lvert \mathbf{a} \rvert\rvert = \sqrt{\mathbf{a} \bullet \mathbf{a}}
$$

<br />


### Vector Direction

A vector's direction is often expressed as the measure of the angle between the horizontal reference axis and the vector.^[While it is common to express the direction as the measure of the angle between the horizontal reference axis, it could also be expressed as the measure of the angle between the vector and any of the reference axes. In more than two-dimensions, the direction would need to include multiple angles to specify the direction; you would need $n-1$ angles to specify an *n*-dimensional vector.] Figure 4 shows this angle (denoted $\theta$) for vector **a**.

```{r fig04, echo=FALSE, fig.width=6, fig.height = 6, out.width='80%', fig.cap="Plot showing vector **a** (in red) in the R1--R2 dimensional space. The direction of this vector is the measure of the angle ($\\theta$) between the horizontal reference axis and the vector."}
graph_paper +
  geom_curve(x = 1.25, y = 0, xend = 0.8, yend = 1.2, curvature = 0.3, color = "#48556C") +
  geom_segment(x = 0, xend = 2, y = 0, yend = 3, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4) +
  annotate("text", x = 1.5, y = 0.9, label = "theta", size = 5, color = "#48556C", parse = TRUE) 
```

To determine this angle, we need to use a [trigonometric function](https://en.wikipedia.org/wiki/Trigonometric_functions). For example, here we can use the fact that the cosine of an angle $\theta$ is defined as the length of the side of the right triangle which is adjacent to $\theta$ divided by the length of the hypotenuse of the triangle.^[You could also use the definition for sine or tangent to compute $\theta$.]

$$
\cos (\theta) = \frac{\lvert\lvert\mathrm{Adjacent~Side}\rvert\rvert}{\lvert\lvert\mathrm{Hypotenuse}\rvert\rvert}
$$


In our example,

$$
\cos (\theta) = \frac{2}{\sqrt{13}}
$$

To find $\theta$, we need to calculate the arc-cosine of $\frac{2}{\sqrt{13}}$. We can compute the arc-cosine in R using the `acos()` function. This function will compute the value of $\theta$ in radians. We can also convert this to degrees by multiplying by $\frac{180}{\pi}$ (where $\pi$ is the mathematical constant approximated as 3.14).

```{r}
# Find theta using arccosine (radians)
acos(2 / sqrt(13))

# Find theta using arccosine (degrees)
acos(2 / sqrt(13)) * 180 / pi
```

We always measure the angle counter-clockwise to the reference axis. So if we measure the direction associated with the vector $\begin{bmatrix}-2 \\ -2\end{bmatrix}$, the direction is more than $\pi$ radians (or $180^\circ$).

```{r fig20, echo=FALSE, fig.width=6, fig.height = 6, out.width='80%', fig.cap="Plot showing vector (-2, -2) (in red) in the R1--R2 dimensional space. The direction of this vector is the measure of the angle ($\\theta$) between the vector and the horizontal reference axis measured in the counter-clockwise direction from the axis."}
graph_paper +
  geom_curve(x = 0.5, y = 0, xend = -1, yend = -1, curvature = 1.5, color = "#48556C") +
  geom_segment(x = 0, y = 0, xend = -2, yend = -2, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5)) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4) +
  annotate("text", x = -1.5, y = 0.9, label = "theta", size = 4, color = "#48556C", parse = TRUE) 
```

There are several ways to compute this angle. Remember that when we use the trignometric functions, we are computing the angle in the right triangle formed by the vector and the reference axes (see Figure \@ref(fig:fig21)). 



```{r fig21, echo=FALSE, fig.width=6, fig.height = 6, out.width='80%', fig.cap="Plot showing vector (-2, -2) (in red) in the R1--R2 dimensional space. The direction of this vector is the measure of the angle ($\\theta$) between the vector and the horizontal reference axis measured in the counter-clockwise direction from the axis. Here we have split this angle up into two parts; $\\alpha$ can be computed using a trigonometric function, and $\\beta = \\pi$ radians."}

triangle_data = data.frame(
  x = c(0, -2, -2),
  y = c(0, 0, -2)
)

graph_paper +
  geom_polygon(data = triangle_data, aes(x = x, y = y), fill = "#FF2D21", alpha = 0.2) +
  geom_curve(x = -1, y = 0, xend = -1, yend = -1, curvature = .3, color = "#48556C") +
  geom_curve(x = 1.3, y = 0, xend = -1.3, yend = 0, curvature = 1.2, color = "#48556C") +
  geom_segment(x = 0, y = 0, xend = -2, yend = -2, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5)) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4) +
  annotate("text", x = -1.5, y = -.7, label = "alpha", size = 4, color = "#48556C", parse = TRUE) +
  annotate("text", x = 1, y = 1.8, label = "beta==pi", size = 4, color = "#48556C", parse = TRUE)
```


To find $\theta$ we need to find $\alpha$ using trigonometric functions and then add it to $\beta$. For example,

$$
\begin{split}
\tan (\alpha) &= \frac{\lvert\lvert\mathrm{Opposite~Side}\rvert\rvert}{\lvert\lvert\mathrm{Adjacent~Side}\rvert\rvert} \\[2ex]
&= \frac{2}{2} \\[2ex]
&= 1
\end{split}
$$

Computing the arc-tangent with the `atan()` function we find that $\alpha=0.785$ radians. Adding this to the $\beta$ value which is $\pi$ radians, we get the vector's direction of 3.927 radians (or $225^\circ$).

```{r}
# Compute alpha (in radians)
atan(1)

# Compute theta = alpha + beta (in radians)
atan(1) + pi

# Compute theta = alpha + beta (in degrees)
(atan(1) + pi) * 180 / pi
```

<br />


## Adding Vectors

Geometrically, adding two vectors, say **a** and **b**, is equivalent to drawing vector **b** so that its tail is placed at the head of vector **a**. Then the sum is the new vector that originates at vector **a**'s tail and terminates at vector **b**'s head. Figure \@ref(fig:fig06) geometrically shows the sum of the following vectors:

$$
\mathbf{a} = \begin{bmatrix}
2 \\ 3
\end{bmatrix} \qquad \mathbf{b} = \begin{bmatrix}
-4 \\ 1
\end{bmatrix} \qquad \mathbf{a}+\mathbf{b} = \begin{bmatrix}
-2 \\ 4
\end{bmatrix}
$$

```{r fig06, echo=FALSE, fig.width=6, fig.height = 6, out.width='60%', fig.cap="Plot showing the sum of vector **a** (in red) and vector **b** (in blue) in the R1--R2 dimensional space. For convenience we have located the tail of vector **a** at the origin. The vector that corresponds to the sum of **a** and **b** is shown in purple."}
graph_paper +
  geom_segment(x = 0, y = 0, xend = 2, yend = 3, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  geom_segment(x = 2, y = 3, xend = -2, yend = 4, color = "#1A5C71",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  geom_segment(x = 0, y = 0, xend = -2, yend = 4, color = "#C1A7C5",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4)  +
  annotate("text", x = 1, y = 1, label = "(2, 3)", hjust = 0, size = 4, color = "#FF2D21") +
  annotate("text", x = 0.75, y = 4.25, label = "(-4, 1)", vjust = 1, size = 4, color = "#1A5C71") +
  annotate("text", x = -2, y = 2, label = "(-2, 4)", vjust = 1, size = 4, color = "#C1A7C5")
```

To subtract vector **b** from vector **a**, we re-write this as vector addition:

$$
\mathbf{a} - \mathbf{b} = \mathbf{a} + (-1)\mathbf{b}
$$

Thus, we can invert the signs on all of the elements of **b** and add this to **a**. 

$$
\begin{bmatrix}
2 \\ 3
\end{bmatrix} + \begin{bmatrix}
4 \\ -1
\end{bmatrix} = \begin{bmatrix}
6 \\ 2
\end{bmatrix}
$$

This is geometrically shown in Figure \@ref(fig:fig07).

```{r fig07, echo=FALSE, fig.width=6, fig.height = 6, out.width='60%', fig.cap="Plot showing the sum of vector **a** (in red) and the inverted vector **b** (in blue) in the R1--R2 dimensional space. For convenience we have located the tail of vector **a** at the origin. The vector that corresponds to the difference of **a** and **b** is shown in purple."}
graph_paper +
  geom_segment(x = 0, y = 0, xend = 2, yend = 3, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  geom_segment(x = 2, y = 3, xend = 6, yend = 2, color = "#1A5C71",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  geom_segment(x = 0, y = 0, xend = 6, yend = 2, color = "#C1A7C5",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4)   +
  annotate("text", x = 0, y = 2, label = "(2, 3)", size = 4, color = "#FF2D21") +
  annotate("text", x = 3.75, y = 3.25, label = "(4, -1)", size = 4, color = "#1A5C71") +
  annotate("text", x = 5, y = 1, label = "(6, 2)", vjust = 1, size = 4, color = "#C1A7C5")
```
<br />

## Scalar Multiplication

Consider the product $\lambda \mathbf{a}$ where,

$$
\lambda = 2 \quad \mathrm{and} \quad \mathbf{a} = \begin{bmatrix}
2 \\ 3
\end{bmatrix}
$$

The product is

$$
\begin{bmatrix}
4 \\ 6
\end{bmatrix}
$$

To show what happens geometrically when we multiply a vector by a scalar, let's examine a plot of the original vector and the product.

```{r fig08, echo=FALSE, fig.width=8, fig.height = 4, out.width='90%', fig.cap="LEFT: Plot showing vector **a** in the R1--R2 dimensional space. RIGHT: Plot showing vector 2(**a**) in the R1--R2 dimensional space. For convenience we have located the tail of both vectors at the origin."}
p1 = graph_paper +
  geom_segment(x = 0, y = 0, xend = 2, yend = 3, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4)

p2 = graph_paper +
  geom_segment(x = 0, y = 0, xend = 4, yend = 6, color = "#1A5C71",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4)

# Display
p1 | p2
```

Multiplying vector **a** by 2 doubled its length. However the direction of the new vector is the same as the original. In general, multiplying a vector by a positive scalar changes the length of a vector but not the direction. If it is multiplied by a negative scalar, not only does the length of the resulting vector change, but its direction is $180^\circ$ from the original. For example, Figure \@ref(fig:fig09) shows vector **a** and the vector $\lambda \mathbf{a}$ where $\lambda = -0.5$.

```{r fig09, echo=FALSE, fig.width=8, fig.height = 4, out.width='90%', fig.cap="LEFT: Plot showing vector **a** in the R1--R2 dimensional space. RIGHT: Plot showing vector -0.5(**a**) in the R1--R2 dimensional space. For convenience we have located the tail of both vectors at the origin."}
p1 = graph_paper +
  geom_segment(x = 0, y = 0, xend = 2, yend = 3, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4)

p2 = graph_paper +
  geom_segment(x = 0, y = 0, xend = -1, yend = -1.5, color = "#1A5C71",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4)

# Display
p1 | p2
```

The product vector is half the length of the original and is pointed in the complete opposite direction.

<br />


:::exercises
### Exercises



<!-- Students' scores on two exams ranged from 32 to 98 out of a possible 100 points on both exams. The scores for six students are shown below for Exam 1 (**x**) and Exam 2 (**y**). -->

<!-- $$ -->
<!-- \mathbf{x} = \begin{bmatrix}56 \\ 64 \\ 32 \\ 88 \\ 90 \\ 79\end{bmatrix} \qquad \mathbf{y} = \begin{bmatrix}50 \\ 69 \\ 51 \\ 98 \\ 87 \\ 70\end{bmatrix} -->
<!-- $$ -->


<!-- 4. Determine the total score for each student by finding $\mathbf{z} = \mathbf{x} + \mathbf{y}$. -->

<!-- <button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_04');">Show/Hide Solution</button> -->
<!-- <div id="solution_04" style="display:none; margin: -40px 0 40px 40px;">$\mathbf{z} = \begin{bmatrix}56 + 50 \\ 64 + 69 \\ 32 + 51 \\ 88 + 98 \\ 90 + 87 \\ 79 + 70\end{bmatrix} = \begin{bmatrix}106 \\ 133 \\ 83 \\ 186 \\ 177 \\ 149\end{bmatrix}$</div> -->


<!-- 5. Determine the mean score for each student by finding $\frac{1}{2} \mathbf{z}$. -->

<!-- <button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_05');">Show/Hide Solution</button> -->
<!-- <div id="solution_05" style="display:none; margin: -40px 0 40px 40px;">$\frac{1}{2}\mathbf{z} = \frac{1}{2}\begin{bmatrix}106 \\ 133 \\ 83 \\ 186 \\ 177 \\ 149\end{bmatrix}= \begin{bmatrix}53.0 \\ 66.5 \\ 41.5 \\ 93.0 \\ 88.5 \\ 74.5\end{bmatrix}$</div> -->

<!-- 6. If the students took a third exam, and the scores for the same six students were presented in vector **w**, how would you write the algebraic expression (using vector notation) to obtain the students' mean scores ?   -->

<!-- <button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_06');">Show/Hide Solution</button> -->
<!-- <div id="solution_06" style="display:none; margin: -40px 0 40px 40px;">$\frac{1}{3}(\mathbf{x} + \mathbf{y} + \mathbf{w})$</div> -->

<!-- 7. If the mean score on Exam 3 was 60, write the algebraic expression (using vector notation) to find the six students' mean deviation scores on Exam 3. -->

<!-- <button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_07');">Show/Hide Solution</button> -->
<!-- <div id="solution_07" style="display:none; margin: -40px 0 40px 40px;">$\mathbf{w} - \begin{bmatrix}60 \\ 60 \\ 60 \\ 60 \\ 60 \\ 60\end{bmatrix}$</div> -->




<!-- Consider the following vectors: -->

<!-- $$ -->
<!-- \mathbf{a} = \begin{bmatrix}1 \\ 2 \\ 4 \end{bmatrix} \qquad \mathbf{b} = \begin{bmatrix}2 \\ 1 \\ -1\end{bmatrix} \qquad \mathbf{c} = \begin{bmatrix}5 \\ 2 \\ 3 \end{bmatrix} \qquad \mathbf{d} = \begin{bmatrix}2 \\ 1 \\ 0 \end{bmatrix} \qquad \mathbf{e} = \begin{bmatrix}1 \\ 3 \\ 2 \end{bmatrix} -->
<!-- $$ -->

<!-- 8. Find $\mathbf{a}\bullet\mathbf{b}$. -->

<!-- <button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_08');">Show/Hide Solution</button> -->
<!-- <div id="solution_08" style="display:none; margin: -40px 0 40px 40px;">$1(2) + 2(1) + 4(-1) = 0$</div> -->


<!-- 9. Find $\mathbf{c}\bullet\mathbf{c}$. -->

<!-- <button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_09');">Show/Hide Solution</button> -->
<!-- <div id="solution_09" style="display:none; margin: -40px 0 40px 40px;">$5(5) + 2(2) + 3(3) = 38$</div> -->


<!-- 10. Find $\mathbf{d}^{\intercal}\mathbf{e}$ and $\mathbf{e}^{\prime}\mathbf{d}$ -->

<!-- <button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_10');">Show/Hide Solution</button> -->
<!-- <div id="solution_10" style="display:none; margin: -40px 0 40px 40px;">$5(5) + 2(2) + 3(3) = 38$</div> -->




**Create a 2-dimensional orthogonal reference system. Using the previously defined vectors, show the following vector operations on that system.**

11. $\mathbf{x}+\mathbf{y}$

<button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_11');">Show/Hide Solution</button>
<div id="solution_11" style="display:none; margin: -40px 0 40px 40px;">

$$
\begin{split}
\mathbf{x} + \mathbf{y} &= \begin{bmatrix} 2 \\ -1\end{bmatrix} + \begin{bmatrix} -3 \\ 2\end{bmatrix} \\[2ex]
&= \begin{bmatrix} -1 \\ 1\end{bmatrix}
\end{split}
$$

```{r echo=FALSE, out.width="60%"}
graph_paper +
  geom_segment(x = 0, y = 0, xend = 2, yend = -1, color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  geom_segment(x = 2, y = -1, xend = -1, yend = 1, color = "#1A5C71",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  geom_segment(x = 0, y = 0, xend = -1, yend = 1, color = "#C1A7C5",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4) +
  coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5)) +
  annotate("text", x = 1, y = -1, label = "x", fontface = "bold", color = "#FF2D21") +
  annotate("text", x = 0.5, y = 0.5, label = "y", fontface = "bold", color = "#1A5C71") +
  annotate("text", x = -1, y = 0.4, label = "x + y", fontface = "bold", color = "#C1A7C5")
```
</div>


12. $3\mathbf{z}$

<button class="solution-btn solution-btn-default" onclick="toggle_visibility('solution_12');">Show/Hide Solution</button>
<div id="solution_12" style="display:none; margin: -40px 0 40px 40px;">
```{r echo=FALSE, out.width="90%", fig.height=4, fig.width=8}
p1 = graph_paper +
  geom_segment(x = 0, y = 0, xend = (1/sqrt(2)), yend = (1/sqrt(2)), color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4) +
  coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5)) +
  annotate("text", x = 2.5, y = 2.5, label = "z", fontface = "bold", color = "#FF2D21")

p2 = graph_paper +
  geom_segment(x = 0, y = 0, xend = 3*(1/sqrt(2)), yend = 3*(1/sqrt(2)), color = "#FF2D21",
    arrow = arrow(length = unit(0.02, "npc"), end = "last", type = "closed")
  ) +
  annotate("text", x = 13.5, y = -0.5, label = "R1", size = 4) +
  annotate("text", x = 0.65, y = 13.5, label = "R2", size = 4) +
  coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5)) +
  annotate("text", x = 2.5, y = 2.5, label = "3z", fontface = "bold", color = "#FF2D21")

# Display
p1 | p2
```
</div>


:::


<br />



