# Cholesky Decompostion

Another decomposition method is Cholesky (or Choleski) decomposition. The Cholesky decomposition method---used in statistical applications from nonlinear optimization, to Monte Carlo simulation methods, to Kalman filtering---is much more computationally efficient than the LU method. The Cholesky method decomposes a symmetric, positive definite matrix **A** into the product of two matrices, $\mathbf{L}$ (a lower-triangular matrix) and $\mathbf{L}^*$ (the *conjugate transpose* of $\mathbf{L}$).^[There are also variations on the Cholesky decomposition method, including LDL and LDL$^\intercal$ decomposition. In these methods, **D** is a diagonal matrix.] 

$$
\underset{n \times n}{\mathbf{A}} = \mathbf{LL}^*
$$

The conjugate transpose is computed by taking the transpose of a matrix and then finding the *complex conjugate* of each element in the matrix. To understand what a complex conjugate is, we first remind you of the idea of a complex number. Remember that all real and imaginary numbers are complex numbers which can be expressed as $a+bi$, where *a* is the real part of the number and *b* is the imaginary part of the number, and *i* is the square root of $-1$. For example the number 2 can be expressed as $2 + 0i$. *Note:* The complex conjugate of a real number is just the real number, since $a+0i = a-0i=a$.

The complex conjugate of a number is itself a complex number that has the exact same real part and an imaginary part equal in magnitude but opposite in sign. For example the complex conjugate for the number $3 + 2i$ is $3 - 2i$. 

As an example, say that matrix **A** was

$$
\mathbf{A} = \begin{bmatrix}
1 & 3+i & -2+3i  \\
0 & 4 & 0-i  \\
\end{bmatrix}
$$

The conjugate transpose of **A**, symbolized as $\mathbf{A}^*$, can be found by first computing the transpose of **A**, and then finding the complex conjugate of each element in the transpose.

$$
\mathbf{A}^{\intercal} = \begin{bmatrix}
1 & 0  \\
3+i0 & 4  \\
-2+3i & 0-i
\end{bmatrix}
$$

And,

$$
\mathbf{A}^{*} = \begin{bmatrix}
1 & 0  \\
3-i0 & 4  \\
-2-3i & 0+i
\end{bmatrix}
$$

:::fyi
Many texts just present the notation for the Cholesky decomposition as $\mathbf{A}=\mathbf{LL}^\intercal$, rather than $\mathbf{A}=\mathbf{LL}^*$. This is because they make the further assumption that all the elements in **A** are real numbers. In that case, $\mathbf{L}^*=\mathbf{L}^\intercal$. We will make that assumption and use that notation moving forward.
:::

<br />

## Example of Cholesky Decomposition

Say we wanted to compute a Cholsky decomposition on a $2 \times 2$ symmetric matrix:

$$
\underset{2\times 2}{\mathbf{A}} = \begin{bmatrix}
5 & -4  \\
-4 & 5  \\
\end{bmatrix}
$$

The Cholesky decomposition would factor **A** into the following:

$$
\begin{split}
\underset{2\times 2}{\mathbf{A}} &= \underset{2\times 2}{\mathbf{L}}~\underset{2\times 2}{\mathbf{L}^\intercal}\\[1em]
\begin{bmatrix}
5 & -4  \\
-4 & 5  \\
\end{bmatrix} &= \begin{bmatrix}
l_{11} & 0  \\
l_{21} & l_{22}  \\
\end{bmatrix} \begin{bmatrix}
l_{11} & l_{21}  \\
0 & l_{22}  \\
\end{bmatrix}
\end{split}
$$

Carrying out the matrix algebra we have the following three unique equations:


$$
\begin{split}
5 &= l_{11}(l_{11}) \\[0.5em]
-4 &= l_{11}(l_{21}) \\[0.5em]
5 &= l_{21}(l_{21}) + l_{22}(l_{22})
\end{split}
$$

Since we have three equations with three unknowns we can solve for each element in **L**.


$$
\begin{split}
l_{11} &= \sqrt{5} \approx 2.24\\[0.5em]
l_{21} &= \dfrac{-4}{\sqrt{5}} \approx -1.79 \\[0.5em]
l_{22} &= \sqrt{1.8} \approx 1.34
\end{split}
$$

So the Cholsky decomposition is,

$$
\begin{bmatrix}
5 & -4  \\
-4 & 5  \\
\end{bmatrix} = \begin{bmatrix}
\sqrt{5}  & 0  \\
\dfrac{-4}{\sqrt{5}} & \sqrt{1.8}  \\
\end{bmatrix} \begin{bmatrix}
\sqrt{5}  & \dfrac{-4}{\sqrt{5}}  \\
0 & \sqrt{1.8}  \\
\end{bmatrix}
$$

Or using the approximations,

$$
\begin{bmatrix}
5 & -4  \\
-4 & 5  \\
\end{bmatrix} \approx \begin{bmatrix}
2.24  & 0  \\
-1.79 & 1.34  \\
\end{bmatrix} \begin{bmatrix}
2.24  & -1.79  \\
0 & 1.34  \\
\end{bmatrix}
$$


<br />

## Cholesky Decomposition using R

We can use the `chol()` function to compute the Cholesky decomposition. For example to carry out the Cholesky decomposition on **A** form the previous section, we would use the following syntax: 

```{r}
# Create A
A = matrix(
  data = c(5, -4, -4, 5), 
  nrow = 2
  )

# Cholesky decomposition
cholesky_decomp = chol(A)

# View results
cholesky_decomp
```

Note that the output from `chol()` is actually the transpose matrix, $\mathbf{L}^\intercal$. To obtain **L**, we need to take the transpose of this output. We can also verify that the decomposition worked.

```{r}
# Obtain L
t(cholesky_decomp)

# Check results L(L^T) = A
t(cholesky_decomp) %*% cholesky_decomp
```

:::protip
`r emo::ji("construction")` CAUTION: The `chol()` function does not check for symmetry. If you use the function to decompose a nonsymmetric matrix, the results will likely be meaningless. 
:::

For example, consider the following nonsymmetric matrix:

$$
\mathbf{A} = \begin{bmatrix}5 & 1\\ -4 & 2\end{bmatrix}
$$

Computing the decomposition, we get:

```{r}
# Create A
A = matrix(
  data = c(5, -4, 1, 2), 
  nrow = 2
  )

# Cholsky decomposition
chol(A)
```

Checking the matrix multiplication we do not get back to the original matrix, **A**:

```{r}
# Check results
t(chol(A)) %*% chol(A)
```

<br />



## Statistical Application: Estimating Regression Coefficents with Cholesky Decomposition

We can again use the the OLS solution to compute the elements of **b** by  solving the system of equations represented in:

$$
(\mathbf{X}^{\intercal}\mathbf{X})\mathbf{b} = \mathbf{X}^{\intercal}\mathbf{y}
$$

So long as $\mathbf{X}^\intercal\mathbf{X}$ is positive definite, we can use Cholesky decomposition to solve for the elements of **b** using the same two-step process we did for LU decomposition. Namely,

$$
\begin{split}
&\mathrm{Solve~~}\mathbf{LZ} &= \mathbf{X}^{\intercal}\mathbf{y}~~\mathrm{for~}\mathbf{Z} \\[2ex]
&\mathrm{Solve~~}\mathbf{L}^\intercal\mathbf{b} & =\mathbf{Z}~~\mathrm{for~}\mathbf{b} 
\end{split}
$$

Using the same simulated data as the example from last chapter, consider the following simulated data set of $n=50$ cases:

```{r}
# Number of cases
n = 50

# Create 50 x-values evenly spread b/w 1 and 500 
x = seq(from = 1, to = 500, len = n)

# Create X matrix
X = cbind(1, x, x^2, x^3)

# Create b matrix
b = matrix(
  data = c(1, 1, 1, 1), 
  nrow = 4
  )

# Create vector of y-values
set.seed(1)
y = X %*% b + rnorm(n, mean = 0, sd = 1)
```

We can again use `forwardsolve()` and `backsolve()` to help solve the two equations.

```{r}
# Obtain L^T and L
L_T = chol(t(X)%*%X)
L = t(L_T)

# Solve for Z
Z = forwardsolve(L,  crossprod(X,y))

# Solve for b
b = backsolve(L_T, Z)

# View b
b
```




<br />
