---
title: "Scalars and Vectors"
output: 
  learnr::tutorial:
    progressive: false
    allow_skip: true
    theme: flatly
    css: ['css/style.css', 'css/notes.css']
    highlight: espresso
    ace_theme: idle_fingers
bibliography: ['references.bib']
csl: 'style/apa-single-spaced.csl'
runtime: shiny_prerendered
---



<!-- Add JavaScript code for making the exercise code larger -->
<script language="JavaScript" src="js/exercise-font-size.js"></script>

```{r setup, include=FALSE}
# load packages
library(learnr)     # rstudio/learnr@bdb2c52
#library(gradethis)  # rstudio-education/gradethis@b2d9ef5
#library(learnrhash) # rundel/learnrhash@714511c
#library(dsbox)      # rstudio-education/dsbox@827bd71
library(thematic)   # rstudio/thematic@0a6b4b7
library(tidyverse)
library(showtext)

# dark theme and larger font size for plots
thematic_on(
  bg = "#212121",
  fg = "#ffffff",
  accent = "#354b63",
  font = font_spec(scale = 1.5, families = "Neucha")
)

# make fonts work
# https://stackoverflow.com/a/55159245/5157940
dir.create("~/.fonts")
file.copy("www/neucha.ttf", "~/.fonts")
system("fc-cache -f ~/.fonts")

# hide non-exercise code chunks
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

In this tutorial, you will learn basic ideas and properties of matrices and vectors. You will also learn about some special matrices (e.g., square matrices). Finally, you will learn how to use R to create matrices and vectors.


<br />

## Scalars

Real numbers are called *scalars*, and *scalar arithmetic* is the arithmetic operations (addition, subtraction, multiplication, and division) we perform using real numbers. For example,

$$
\begin{split}
2 + 3 = 5 \\[1ex]
21 \div 3 = 7
\end{split}
$$

<br />


## Vectors

A *vector* is a specifically ordered array of two or more values. Each value in a vector is called an *element* or *component*. Vectors are described in terms of the number of elements they have. For example, the following is a two-element vector:

$$
\begin{bmatrix}
5 & 4
\end{bmatrix}
$$

A vector may be written as a row or a column, and are respectively referred to as *row vectors* or *column vectors*. Vectors are typically denoted using a lower-case, underlined letter ($\underline{a}$), a lower-case bold letter (**a**), or a lower-case letter with an overset arrow ($\vec{a}$); different texts and authors use different notation.

For example, here, **a** is a four-element column vector:

$$
\mathbf{a} = \begin{bmatrix}
5 \\ 4 \\ 7 \\2
\end{bmatrix}
$$
Transposition is an operation in which we replace the *i*th element of a column vector is instead entered as the *i*th element of a row vector, and vice-versa. In other words, we are converting a column vector into a row vector, or, conversely, converting a row vector into a column vector. Notationally, we use a superscripted prime to denote a vector's transpose. For example,

$$
\mathbf{a}^{\prime} = \begin{bmatrix}
5 & 4 & 7 & 2
\end{bmatrix}
$$
:::fyi
In this text, we will use a bold-face, lower-case letter to denote a column vector. Row vectors will be notated as a vector transpose.  
:::

The general form of a column vector with *n* elements is:

$$
\mathbf{x} = \begin{bmatrix}
x_1 \\ x_2 \\ x_3 \\ \vdots \\ x_n
\end{bmatrix}
$$

In statistical or psychometric appplications, this is typically how we think of and structure data collected on a particular attribute for many cases. For example, **x** might represent test scores for $n=100$ students. This vector would contain 100 elements, each corresponding to a student's test score. In this case, $x_1$ would be the test score for the first student recorded in the vector, $x_2$ would be the test score for the second student recorded in the vector, etc.  

<br />

### Exercises

Identify each of the following as a row vector, column vector, or scalar.

```{r ex-01, echo=FALSE}
question("$[a_1 + a_2 + a_3 + a_4]$",
  answer("Row Vector", message = "Hint: Summing results in a real number."),
  answer("Column Vector", message = "Try again."),
  answer("Scalar", correct = TRUE)
)
```

```{r ex-02, echo=FALSE}
question("$\\begin{bmatrix} 520 \\\\ 640 \\\\ 780\\end{bmatrix}$",
  answer("Row Vector", message = "Try again."),
  answer("Column Vector", correct = TRUE),
  answer("Scalar", message = "Try again.")
)
```

```{r ex-03, echo=FALSE}
question("13",
  answer("Row Vector", message = "Try again."),
  answer("Column Vector", message = "Try again."),
  answer("Scalar", correct = TRUE)
)
```

```{r ex-04, echo=FALSE}
question("$\\begin{bmatrix} 115 & 129 & 92 & 89\\end{bmatrix}$",
  answer("Row Vector", correct = TRUE),
  answer("Column Vector", message = "Try again."),
  answer("Scalar", message = "Try again.")
)
```

```{r ex-05, echo=FALSE}
question("$5 \\times 11$",
  answer("Row Vector", message = "Try again."),
  answer("Column Vector", message = "Try again."),
  answer("Scalar", correct = TRUE)
)
```

```{r ex-06, echo=FALSE}
question("$\\begin{bmatrix} 0 & 0 & 0\\end{bmatrix}$",
  answer("Row Vector", correct = TRUE),
  answer("Column Vector", message = "Try again."),
  answer("Scalar", message = "Try again.")
)
```

<br />


### Vector Equality

Vectors may be equal to other vectors if they satisfy three conditions:

1. Same number of elements;
2. Corresponding elements from each vector must be equal; and
3. Both vectors must be row-vectors or both must be column-vectors

<br />


#### Exercises

Indicate whether the vectors are equal or not equal.

```{r ex-07, echo=FALSE}
question("$\\begin{bmatrix} 5 \\\\ 6 \\end{bmatrix} \\overset{?}{=} \\begin{bmatrix} 6 \\\\ 5 \\end{bmatrix}$",
  answer("Equal", message = "Are the corresponding elements from both vectors equal?"),
  answer("Not equal", correct = TRUE)
)
```

```{r ex-08, echo=FALSE}
question("$\\begin{bmatrix} 9 \\\\ 0 \\\\ 1 \\end{bmatrix} \\overset{?}{=} \\begin{bmatrix} 9 \\\\ 0 \\\\ 1 \\end{bmatrix}$",
  answer("Equal", correct = TRUE),
  answer("Not equal", message = "Are all three conditions for vector equality met?")
)
```

```{r ex-09, echo=FALSE}
question("$\\begin{bmatrix} 55 \\\\ 66 \\\\ 48 \\end{bmatrix} \\overset{?}{=} \\begin{bmatrix} 55 & 66 & 48 \\end{bmatrix}$",
  answer("Equal", message = "Are both vectors row vectors? Are they both column vectors?"),
  answer("Not equal", correct = TRUE)
)
```

<br />


## Vector Operations

Some arithmetic operations (addition, subtraction, and multiplication) can be performed with vectors. However, how these operations are carried out depends on whether one is working with all vectors or with a vector and a scalar. We will illustrate some of the common vector operations in this section.

<br />


### Vector Addition and Subtraction

Vectors can be added together if they:

1.	Have the same number of elements; and
2.	Are both row vectors or both column vectors.

To add two column vectors (or two row vectors) together, each of length *n*, we sum the corresponding elements in the vectors. For example, consider the column vectors **a** and **b**, where,

$$
\mathbf{a} = \begin{bmatrix}5 \\ 2\end{bmatrix} \qquad \mathbf{b} = \begin{bmatrix}1 \\ 3\end{bmatrix}
$$

We can compute the sum of these vectors as:

$$
\begin{split}
\mathbf{a} + \mathbf{b}  &= \begin{bmatrix}5 \\ 2\end{bmatrix} + \begin{bmatrix}1 \\ 3\end{bmatrix} \\[1ex]
&= \begin{bmatrix}5 + 1 \\ 2 + 3\end{bmatrix} \\[1ex]
&= \begin{bmatrix}6 \\ 5\end{bmatrix}
\end{split}
$$
In general, 

$$
\begin{split}
\mathbf{a} + \mathbf{b}  &= \begin{bmatrix}a_1 \\ a_2 \\ a_3 \\ \vdots \\ a_n\end{bmatrix} + \begin{bmatrix}b_1 \\ b_2 \\ b_3 \\ \vdots \\ b_n\end{bmatrix} \\[1ex]
&= \begin{bmatrix}a_1 + b_1 \\ a_2 + b_2 \\ a_3 + b_3 \\ \vdots \\ a_n + b_n\end{bmatrix}
\end{split}
$$

In vector addition, since each corresponding element is added, vector addition satisfies both the commutative and associative properties. That is,

$$
\mathbf{a} + \mathbf{b} = \mathbf{b} + \mathbf{a}
$$
and

$$
\begin{split}
\mathbf{a} + (\mathbf{b} + \mathbf{c}) &= (\mathbf{a} + \mathbf{b}) + \mathbf{c} \\[1ex]
&= \mathbf{a} + \mathbf{b} + \mathbf{c}
\end{split}
$$
Convince yourself these two properties are satisfied using the following vectors.

$$
\mathbf{a} = \begin{bmatrix}5 \\ 2\end{bmatrix} \qquad \mathbf{b} = \begin{bmatrix}1 \\ 3\end{bmatrix} \qquad \mathbf{c} = \begin{bmatrix}2 \\ 0\end{bmatrix}
$$

These same conditions and properties also apply for vector subtraction. For example,

$$
\begin{split}
\mathbf{a} - \mathbf{b}  &= \begin{bmatrix}5 \\ 2\end{bmatrix} - \begin{bmatrix}1 \\ 3\end{bmatrix} \\[1ex]
&= \begin{bmatrix}5 - 1 \\ 2 - 3\end{bmatrix} \\[1ex]
&= \begin{bmatrix}4 \\ -1\end{bmatrix}
\end{split}
$$

<br />


### Multiplication of Vectors by a Scalar

When a vector is multiplied by a scalar, each element of the vector is multiplied by the value of the scalar. Consider the following vector **a** and scalar $\lambda$,

$$
\mathbf{a} = \begin{bmatrix}2 \\ 3 \\ 0 \\ -1\end{bmatrix} \qquad \lambda=3
$$

Multiplying  **a** by $\lambda$ gives:

$$
\begin{split}
\lambda\mathbf{a} &= 3\begin{bmatrix}2 \\ 3 \\ 0 \\ -1\end{bmatrix} \\[1ex]
&= \begin{bmatrix}3 \times 2 \\ 3 \times 3 \\ 3 \times 0 \\ 3 \times -1\end{bmatrix} \\[1ex]
&= \begin{bmatrix}6 \\ 9 \\ 0 \\ -3\end{bmatrix}
\end{split}
$$

This same method applies to row vectors. Scalars may be fractions, negative numbers, or unknowns. For example, if the scalar $\gamma$ is an unknown scalar, then

$$
\gamma\mathbf{a} = \begin{bmatrix}2\gamma \\ 3\gamma \\ 0 \\ -1\gamma\end{bmatrix}
$$

<br />


### Division by a Scalar

Technically, division is undefined for vectors. However, if we are dividing by a scalar, we can multiply the vector by the scalar's reciprocal. For example, using our previously defined vector **a** and scalar $\lambda$,

$$
\mathbf{a} \div \lambda \quad \mathrm{does~not~exist}
$$

But, we can multiply by the reprical of $\lambda$, namely $\dfrac{1}{3}$.

$$
\begin{split}
\lambda\mathbf{a} &= \frac{1}{3}\begin{bmatrix}2 \\ 3 \\ 0 \\ -1\end{bmatrix} \\[1ex]
&= \begin{bmatrix} \frac{2}{3} \\ 1 \\ 0 \\ -\frac{1}{3}\end{bmatrix}
\end{split}
$$

### Exercises

Students' scores on a midterm and final ranged from 32 to 98 out of a possible 100 points on both exams. The scores for six students are shown below for the midterm (**x**) and final (**y**).

$$
\mathbf{x} = \begin{bmatrix}56 \\ 64 \\ 32 \\ 88 \\ 90 \\ 79\end{bmatrix} \qquad \mathbf{y} = \begin{bmatrix}50 \\ 69 \\ 51 \\ 98 \\ 87 \\ 70\end{bmatrix}
$$

Consider finding the total scores for each student by computing **z** = **x** + **y**.


<!-- ```{r, echo=FALSE} -->
<!-- library(shinyFeedback) -->
<!-- useShinyFeedback() -->

<!-- numericInput("num_elm", "How many elements would <b>z</b> include?", value = 0) -->
<!-- verbatimTextOutput("value") -->
<!-- ``` -->

<!-- ```{r, context="server"} -->
<!-- observeEvent(input$num_elm, { -->

<!--     if (is.na(input$num_elm)) { -->
<!--       hideFeedback("num_elm") -->
<!--       } else if (input$num_elm != 6 & !is.na(input$num_elm)) { -->
<!--       showFeedbackWarning( -->
<!--         inputId = "num_elm", -->
<!--         text = "Not quite. Try again." -->
<!--       )   -->
<!--     } else if (input$num_elm == 6) { -->
<!--       showFeedbackSuccess( -->
<!--         inputId = "num_elm", -->
<!--         text = "Yes! Good job!." -->
<!--       )   -->
<!--     } -->

<!--   }) -->
<!-- ``` -->

1. What is the total score for each student?  Find $\mathbf{z} = \mathbf{x} + \mathbf{y}$.

2. What is the mean score for each student? To determine this, find $\frac{1}{2} \mathbf{z}$.

3. If the students took three exams, and the scores for the same six students were presented in vector **w**, how would you write the computation to obtain the students' mean scores in vector notation?  

4. If the mean score on the final was 60, find the six students' mean deviation scores employing vector notation.

<br />


### Multiplication of Vectors

There are two types of vector multiplication, the *inner product* and the *outer product*. The inner product yields a scalar value while the outer product yields a rectangular array of values.

To find the inner product (i.e., scalar product, dot product), we multiply corresponding elements of each vector and add the products together. Consider the vectors,


$$
\mathbf{a} = \begin{bmatrix}5 \\ 4 \\ 7 \\ 2\end{bmatrix} \qquad \mathbf{b} = \begin{bmatrix}1 \\ 0 \\ -1 \\ 2\end{bmatrix}
$$

The inner product, or $\mathbf{a} \bullet \mathbf{b}$, is:

$$
\begin{split}
\mathbf{a} \bullet \mathbf{b} &= 5(1) + 4(0) + 7(-1) + 2(2) \\[1ex]
&= 2
\end{split}
$$
Technically, the inner product is the product of a row vector and column vector. In our example this is:

$$
\begin{split}
\mathbf{a} \bullet \mathbf{b} &= \mathbf{a}^{\prime}\mathbf{b} \\[1ex]
&= \begin{bmatrix}5 & 4 & 7 & 2\end{bmatrix} \begin{bmatrix}1 \\ 0 \\ -1 \\ 2\end{bmatrix}
\end{split}
$$

In general, the inner product of two vectors, **a** and **b**, each having *n* elements is:

$$
\begin{split}
\mathbf{a} \bullet \mathbf{b} &= \mathbf{a}^{\prime}\mathbf{b} \\[1ex]
&= \begin{bmatrix}a_1 & a_2 & a_3 & \dots & a_n\end{bmatrix} \begin{bmatrix}b_1 \\ b_2 \\ b_3 \\ \vdots \\ b_n\end{bmatrix} \\[1ex]
&= a_1(b_1) + a_2(b_2) + a_3(b_3) + \ldots + a_n(b_n) \\[1ex]
&= \sum_{i=1}^n a_ib_i
\end{split}
$$
Some caveats:

- The inner product $\mathbf{a}\bullet\mathbf{b}$ is always the product of $\mathbf{a}^{\prime}\mathbf{b}$. (The row vector is always first in the product;  $\mathbf{a}\mathbf{b}^{\prime} \neq \mathbf{a}^{\prime}\mathbf{b}$.)
- It is not possible to find the inner product of more than two vectors. 

<br />


#### Exercises

Consider the following vectors:

$$
\mathbf{a} = \begin{bmatrix}1 \\ 2 \\ 4 \end{bmatrix} \qquad \mathbf{b} = \begin{bmatrix}2 \\ 1 \\ -1\end{bmatrix} \qquad \mathbf{c} = \begin{bmatrix}5 \\ 2 \\ 3 \end{bmatrix} \qquad \mathbf{d} = \begin{bmatrix}2 \\ 1 \\ 0 \end{bmatrix} \qquad \mathbf{e} = \begin{bmatrix}1 \\ 3 \\ 2 \end{bmatrix}
$$

1. Find $\mathbf{a}\bullet\mathbf{b}$.

2. Find $\mathbf{c}\bullet\mathbf{c}$.

3. Find $\mathbf{d}^{\prime}\mathbf{e}$ and $\mathbf{e}^{\prime}\mathbf{d}$

<br />


### Special Vectors

<!-- Vectors are matrices that have a single row or a single column. They are referred to as *row vectors* or *column vectors*, respectively. Below, **A** is a row vector and **B** is a column vector.  -->

<!-- $$ -->
<!-- \underset{1\times 4}{\mathbf{A}} = \begin{bmatrix} -->
<!-- 4 & 1 & -2 & 5 -->
<!-- \end{bmatrix} \qquad \underset{4\times 1}{\mathbf{B}} = \begin{bmatrix} -->
<!-- 4 \\  -->
<!-- 1 \\ -->
<!-- -2 \\ -->
<!-- 5 -->
<!-- \end{bmatrix} -->
<!-- $$ -->

<!-- <br /> -->



<!-- ## Matrices -->

<!-- A matrix is a rectangular array of elements arranged in rows and columns. We typically denote matrices using a bold-faced font. For example, consider the matrix **B** which has 3 rows and 2 columns: -->


<!-- $$ -->
<!-- \mathbf{B} = \begin{bmatrix} -->
<!-- 5 & 1 \\ -->
<!-- 7 & 3 \\ -->
<!-- -2 & -1 -->
<!-- \end{bmatrix} -->
<!-- $$ -->

<!-- The number of rows and columns is denoted as $3\times 2$ and is referred to as the *dimensions* or *order* of the matrix. The dimension is often appended to the bottom of the matrix name: -->

<!-- $$ -->
<!-- \underset{3\times 2}{\mathbf{B}} = \begin{bmatrix} -->
<!-- 5 & 1 \\ -->
<!-- 7 & 3 \\ -->
<!-- -2 & -1 -->
<!-- \end{bmatrix} -->
<!-- $$ -->

<!-- The *elements* within the matrix are indexed by their row number and column number (i.e., the element's position). For example, since the element in the second row and first column is 7, we could write -->

<!-- $$ -->
<!-- \mathbf{B}_{2,1} = 7 -->
<!-- $$ -->

<!-- The subscripts on each element indicate the row and column positions of the element. -->

<!-- <br /> -->


<!-- ### General Form of a Matrix -->

<!-- More generally, we define matrix **A**, which has *n* rows and *k* columns as: -->

<!-- $$ -->
<!-- \underset{n\times k}{\mathbf{A}} = \begin{bmatrix} -->
<!-- a_{11} & a_{12} & a_{13} & \ldots & a_{1k} \\  -->
<!-- a_{21} & a_{22} & a_{23} & \ldots & a_{2k} \\ -->
<!-- a_{31} & a_{32} & a_{33} & \ldots & a_{3k} \\ -->
<!-- \vdots & \vdots & \vdots & \vdots & \vdots \\ -->
<!-- a_{n1} & a_{n2} & a_{n3} & \ldots & a_{nk} -->
<!-- \end{bmatrix} -->
<!-- $$ -->

<!-- where element $a_{ij}$ is in the $i^{\mathrm{th}}$ row and $j^{\mathrm{th}}$ column of **A**. -->

<!-- <br /> -->


<!-- ## Square Matrices -->

<!-- When the number of rows and columns are equal ($n=k$), the matrix is referred to as a *square matrix*. For example, **X** and **Y** are both square matrices. -->

<!-- $$ -->
<!-- \underset{2\times 2}{\mathbf{X}} = \begin{bmatrix} -->
<!-- 4 & 1 \\  -->
<!-- 0 & 3 -->
<!-- \end{bmatrix} \qquad \underset{3\times 3}{\mathbf{Y}} = \begin{bmatrix} -->
<!-- 0 & 1 & 0 \\  -->
<!-- 1 & 3 & -8\\ -->
<!-- 10 & 4 & -2 -->
<!-- \end{bmatrix} -->
<!-- $$ -->


<!-- Square matrices have many properties that make them useful in many statistical applications. For example, they are used in Ordinary Least Squares estimation to calculate the coefficients and standard errors for a regression model. They are also used in principal components analysis, and factor analysis. -->

<!-- <br /> -->


<!-- ### Main Diagonal -->

<!-- In a square matrix, the main diagonal includes the elements that lie along the diagonal from the upper-left element to the lower-right element. For example, the main diagonal in **Y** (highlighted in Magenta) includes the elements $0$, $3$, and $-2$. -->

<!-- $$ -->
<!-- \underset{3\times 3}{\mathbf{Y}} = \begin{bmatrix} -->
<!-- \color{Magenta}0 & 1 & 0 \\  -->
<!-- 1 & \color{Magenta}3 & -8\\ -->
<!-- 10 & 4 & \color{Magenta}{-2} -->
<!-- \end{bmatrix} -->
<!-- $$ -->

<!-- In principal components analysis, the main diagonal plays an important role. -->


<!-- <br /> -->


<!-- ### Matrix Trace -->

<!-- The trace of a matrix is the sum of its main diagonal elements. For example, consider the matrix **Y**:   -->

<!-- $$ -->
<!-- \underset{3\times 3}{\mathbf{Y}} = \begin{bmatrix} -->
<!-- \color{Magenta}0 & 1 & 0 \\  -->
<!-- 1 & \color{Magenta}3 & -8\\ -->
<!-- 10 & 4 & \color{Magenta}{-2} -->
<!-- \end{bmatrix} -->
<!-- $$ -->

<!-- The trace of **Y** is 1. -->

<!-- $$ -->
<!-- \begin{split} -->
<!-- \mathrm{tr}(Y) &= 0 + 3 + -2 \\[1ex]  -->
<!-- &= 1 -->
<!-- \end{split} -->
<!-- $$ -->

<!-- The trace of a matrix is used to calculate the effective degrees of freedom for a model.  -->

<!-- <br /> -->


<!-- ## Matrices in R -->

<!-- To enter a matrix in R, use the `matrix()` function. The elements of the matrix will be filled-in by columns. For example to create matrix **B** from the example above, we can use the following syntax. -->

<!-- ```{r echo=TRUE} -->
<!-- # Define B -->
<!-- B = matrix( -->
<!--   data = c(5, 7, -2, 1, 3, -1), -->
<!--   nrow = 3, -->
<!--   ncol = 2 -->
<!--   ) -->

<!-- # Display B -->
<!-- B -->
<!-- ``` -->


<!-- The `byrow=TRUE` argument will fill the elements by rows rather than columns. Only one of the arguments `nrow=` or `ncol=` are needed, as the number of elements in `data=` will define the other argument. -->



<!-- <br /> -->


<!-- The `dim()` function can be used to return the dimensions of a matrix. -->

<!-- ```{r echo=TRUE} -->
<!-- # Get dimensions for B -->
<!-- dim(B) -->
<!-- ``` -->

<!-- <br /> -->



<!-- To create a vector in R, use the `matrix()` function and the argument of `nrow=1` (row vector) or `ncol=1` (column vector). -->

<!-- <br /> -->

<!-- ### Find the Main Diagonal Using R -->

<!-- We can use the `diag()` function to return the elements on the main diagonal in a square matrix. -->

<!-- ```{r} -->
<!-- Y = matrix( -->
<!--   data = c(0, 1, 10, 1, 3, 4, 0, -8, -2), -->
<!--   nrow = 3 -->
<!--   ) -->

<!-- # Display Y -->
<!-- Y -->

<!-- # Find diagonal elements -->
<!-- diag(Y) -->
<!-- ``` -->

<!-- `r emo::ji("construction")` CAUTION: The `diag()` function also works on non-square matrices. However, it returns the elements on the diagonal starting with the element in the first row and column. -->

<!-- ```{r} -->
<!-- X = matrix( -->
<!--   data = c(0, 1, 1, 3, 4, 0), -->
<!--   nrow = 3 -->
<!--   ) -->

<!-- # Display X -->
<!-- X -->

<!-- # Find 'diagonal' elements -->
<!-- diag(X) -->
<!-- ``` -->

<!-- <br /> -->

<!-- #### Computing the Trace in R -->

<!-- To compute the trace we use the `sum()` and `diag()` functions. -->

<!-- ```{r} -->
<!-- # Compute trace of Y -->
<!-- sum(diag(Y)) -->
<!-- ``` -->

<!-- You can also use the `tr()` function from the **psych** library to compute the trace. -->

<!-- ```{r message=FALSE} -->
<!-- library(psych) -->
<!-- tr(Y) -->
<!-- ``` -->



<!-- <br /> -->

<!-- ## Exercises -->




<!-- Use R to create the following matrix: -->

<!-- $$ -->
<!-- \mathbf{X} = \begin{bmatrix} -->
<!-- 1 & -1 & 4 \\ -->
<!-- 1 & 7 & 5 -->
<!-- \end{bmatrix} -->
<!-- $$ -->


<!-- ```{r create-x, exercise=TRUE, exercise.lines = 5} -->
<!-- X = matrix() -->
<!-- ``` -->



<!-- ```{r dim-x, echo=FALSE} -->
<!-- question("What are the dimensions of **X**?", -->
<!--   answer("2x2", message = "Remember the dimensions are: rows x columns."), -->
<!--   answer("2x3", correct = TRUE), -->
<!--   answer("3x2", message = "Remember the dimensions are: rows x columns."), -->
<!--   answer("3x3", message = "Remember the dimensions are: rows x columns.") -->
<!-- ) -->
<!-- ``` -->


<!-- *Here's a simple exercise with an empty code chunk provided for entering the answer.* -->

<!-- Write the R code required to add two plus two: -->

<!-- ```{r two-plus-two, exercise=TRUE} -->

<!-- ``` -->

<!-- ### Exercise with Code -->

<!-- *Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.* -->

<!-- Now write a function that adds any two numbers and then call it: -->

<!-- ```{r add-function, exercise=TRUE, exercise.lines = 5} -->
<!-- add <- function() { -->

<!-- } -->
<!-- ``` -->

<!-- ## Topic 2 -->

<!-- ### Exercise with Hint -->

<!-- *Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.* -->

<!-- Modify the following code to limit the number of rows printed to 5: -->

<!-- ```{r print-limit, exercise=TRUE, exercise.eval=TRUE} -->
<!-- mtcars -->
<!-- ``` -->

<!-- ```{r print-limit-hint} -->
<!-- head(mtcars) -->
<!-- ``` -->

<!-- ### Quiz -->

<!-- *You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.* -->

<!-- Some questions to verify that you understand the purposes of various base and recommended R packages: -->

<!-- ```{r quiz} -->
<!-- quiz( -->
<!--   question("Which package contains functions for installing other R packages?", -->
<!--     answer("base"), -->
<!--     answer("tools"), -->
<!--     answer("utils", correct = TRUE), -->
<!--     answer("codetools") -->
<!--   ), -->
<!--   question("Which of the R packages listed below are used to create plots?", -->
<!--     answer("lattice", correct = TRUE), -->
<!--     answer("tools"), -->
<!--     answer("stats"), -->
<!--     answer("grid", correct = TRUE) -->
<!--   ) -->
<!-- ) -->
<!-- ``` -->

